``Interestingness'' in Mathematical Theory Exploration

### 2019-11-29 PhDThesis/litreview_interestingness.tex
Shrink table to fit
Fix unclosed footnote

% FIXME: This is original work, so should maybe be in a different section
\begin{table}
  \centering
  \begin{tabular}{ |l|l|c|c|c|c|c|c| }
    \hline
    \multirow{2}{*}{\textbf{Program}}                      &
    \multirow{2}{*}{\textbf{Conjecture Types}}             &
    \multicolumn{6}{c|}{\textbf{Interestingness Measures}} \\ \hhline{~~------}
    \tRow{             &                     & \iE & \iN & \iS & \iA & \iC & \iU}
    \tRow{AM           & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
    \tRow{GT           & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
    \tRow{Graffiti     & \tINE               &   X &   X &   X &     &   X &   X}
    \tRow{\Bagai{}     & \tNE                &     &   X &     &   X &   X &    }
    \tRow{HR           & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
    \tRow{\quickspec{} & \tEQ                &   X &   X &     &   X &   X &    }
    \tRow{\speculate{} & \tCON\ \tEQ / \tINE &   X &   X &     &   X &   X &    }
    \tRow{\isacosy{}   & \tEQ                &   X &   X &     &   X &   X &   X}
    \tRow{\isascheme{} & \tEQ                &   X &   X &     &   X &   X &   X}
  \end{tabular}
  \caption{Classification of MTE tools from~\cite{colton2000notion}, extended
    to include four more recent tools. The interestingness measures are
    \iE{}mpirical plausibility, \iN{}ovelty, \iS{}urprisingness,
    \iA{}pplicability, \iC{}omprehensibility (low complexity) and \iU{}tility.}
  \label{table:colton}
\end{table}

@@

When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of
which $S$ is a special case. Although it seems counterintuitive, a generalised
statement can sometimes be \emph{easier} to prove than the original. This arises
often in inductive proofs (e.g. when reasoning about tail-recursive
definitions~\cite{kapur2003automatic}\footnote{Tail-recursive functions are
  desirable due to their constant space usage, unlike recursion in non-tail
  positions which requires a growing number of stack frames or nested closures.
  Reasoning about non-tail recursive definitions is often simpler, since they
  don't expose accumulator arguments, so equivalence ), since the specific obligations which
arise in the proof may be incompatible with the available inductive hypotheses.
An informative example is given by Boyer and Moore of the associativity of
multiplication in ACL2~\cite{boyer1983proof}:

@@

The interestingness of such generated lemmas derives from both the existence of
the original lemma (which is presumably interesting enough to prove and
subsequently explore), that of the target theorem it may help to prove, and the
novelty provided by reframing existing concepts in a new domain. The running
examples for demonstrating \textsc{ACL2(ml)} are equivalence theorems for
tail-recursive and their non-tail-recursive counterparts, as well as the effect
of repeating certain list operations:

\begin{itemize}

  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fact-tail}(n) = \texttt{fact}(n)$ where \texttt{natp} is the predicate that $n$ is a natural number, whilst \texttt{fact-tail} and \texttt{fact} are tail-recursive and non-tail-recursive implementations of factorial, respectively.

@@

  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{int}(x, x) = x$, where \texttt{int} is the intersection of lists (i.e. a list of elements common to each).

\end{itemize}

@@

\begin{haskell}
plus :: Nat -> Nat -> Nat
plus      n  Z    = n
plus      n (S m) = S (plus n m)

plus_tail :: Nat -> Nat -> Nat
plus_tail n  Z    = n
plus_tail n (S m) = plus_tail (S n) m
\end{haskell}

>>>>

% FIXME: This is original work, so should maybe be in a different section
\begin{table}
  \centering
  \scalebox{0.9}{
    \begin{tabular}{ |l|l|c|c|c|c|c|c| }
      \hline
      \multirow{2}{*}{\textbf{Program}}                      &
      \multirow{2}{*}{\textbf{Conjecture Types}}             &
      \multicolumn{6}{c|}{\textbf{Interestingness Measures}} \\ \hhline{~~------}
      \tRow{             &                     & \iE & \iN & \iS & \iA & \iC & \iU}
      \tRow{AM           & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
      \tRow{GT           & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
      \tRow{Graffiti     & \tINE               &   X &   X &   X &     &   X &   X}
      \tRow{\Bagai{}     & \tNE                &     &   X &     &   X &   X &    }
      \tRow{HR           & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
      \tRow{\quickspec{} & \tEQ                &   X &   X &     &   X &   X &    }
      \tRow{\speculate{} & \tCON\ \tEQ / \tINE &   X &   X &     &   X &   X &    }
      \tRow{\isacosy{}   & \tEQ                &   X &   X &     &   X &   X &   X}
      \tRow{\isascheme{} & \tEQ                &   X &   X &     &   X &   X &   X}
    \end{tabular}
  }
  \caption{Classification of MTE tools from~\cite{colton2000notion}, extended
    to include four more recent tools. The interestingness measures are
    \iE{}mpirical plausibility, \iN{}ovelty, \iS{}urprisingness,
    \iA{}pplicability, \iC{}omprehensibility (low complexity) and \iU{}tility.}
  \label{table:colton}
\end{table}

@@

When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of
which $S$ is a special case. Although it seems counterintuitive, a generalised
statement can sometimes be \emph{easier} to prove than the original. This arises
often in inductive proofs (e.g. when reasoning about tail-recursive
definitions~\cite{kapur2003automatic}\footnote{Tail-recursive functions are
  desirable due to their constant space usage, unlike recursion in non-tail
  positions which requires a growing number of stack frames or nested closures.
  The presence of accumulator arguments makes reasoning harder, so it is common
  to use the direct implementation for proofs, then transport them over to the
  tail-recursive definition via an equivalence relation.}), since the specific
obligations which arise in the proof may be incompatible with the available
inductive hypotheses. An informative example is given by Boyer and Moore of the
associativity of multiplication in \textsc{ACL2}~\cite{boyer1983proof}:

@@

The interestingness of such generated lemmas derives from both the existence of
the original lemma (which is presumably interesting enough to prove and
subsequently explore), that of the target theorem it may help to prove, and the
novelty provided by reframing existing concepts in a new domain. The running
examples for demonstrating \textsc{ACL2(ml)} are equivalence theorems for
tail-recursive definitions with their direct (non-tail-recursive) counterparts,
as well as the effect of repeating certain list operations:

\begin{itemize}
\item
  $\forall n, \texttt{natp}(n) \rightarrow \texttt{fact-tail}(n) =
  \texttt{fact}(n)$ where \texttt{natp} is the predicate that $n$ is a natural
  number, whilst \texttt{fact-tail} and \texttt{fact} are tail-recursive and
  non-tail-recursive implementations of factorial, respectively.

@@

\item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{int}(x, x) = x$,
  where \texttt{int} is the intersection of lists (i.e. a list of elements
  common to each).
\end{itemize}

@@

\begin{haskell}
plus :: Nat -> Nat -> Nat
plus      Z      m = m
plus      (S n') m = S (plus n' m)

plus_tail :: Nat -> Nat -> Nat
plus_tail Z      m = m
plus_tail (S n') m = plus_tail n' (S m)
\end{haskell}
